---
title: "dbtest"
output: github_document
---

[![Build Status](https://travis-ci.org/rstudio/dbtest.svg?branch=master)](https://travis-ci.org/rstudio/dbtest)

# Overview

`dbtest` uses `testthat` to automate testing of `dbplyr` translations by running the tests against live database connections.  Tests are written in YAML files, and connections are either passed explicitly, read from system DSNs, or configured in YAML files.  Further, it is possible to execute a test suite against multiple database connections.

# Install

To install `dbtest`, you can install the latest version from GitHub:

```{r eval=FALSE}
remotes::install_github("rstudio/dbtest")
```

# Connection

The first step to use `dbtest` is to set up a [DBI](todo) connection object.  There are many ways you can do this.

### DSN

Sometimes, a system already has a handful of DSNs (Data Source Names) set up that make connection easy.  In usual DBI, the connection might look like `DBI::dbConnect(odbc::odbc(), "My Data Source")`.  On a linux operating system, these are usually defined in `/etc/odbc.ini` or `~/.odbc.ini`.  If you have DSNs defined on your system, you can utilize them by using the value `"dsn"` as your connection.  `dbtest` will use all of your DSNs and execute tests against them.

```{r eval=FALSE}
dbtest::test_databases("dsn")
```

### YAML File

Every database has different connection parameters.  To make database connections easy to automate, `dbtest` will read a YAML file and pass the named parameters into `dbConnect` to create a DBI connection.  Note that the [`config`](todo) package is used, so you must label the set of connections and refer to it with `R_CONFIG_ACTIVE=mylabel`.  Otherwise, the `default` heading will be selected. An example might look like:

```yaml
default:
  pg:
    drv: !expr odbc::odbc()
    Driver: PostgreSQL
    Host: postgres.example.com
    Port: 5432
    Database: postgres
    UID: user
    PWD: password

  mssql:
    drv: !expr odbc::odbc()
    Driver: SQLServer
    Server: mssql.example.com
    Port: 1433
    UID: user
    PWD: password

  oracle:
    drv: !expr odbc::odbc()
    Driver: Oracle
    Host: oracle.example.com
    Port: 1521
    SVC: xe
    UID: user
    PWD: password
```

Notice that the names of the various databases are different.  This corresponds to the parameters that different database providers expect in the `dbConnect` function.  Notice also that `drv` is `!expr odbc::odbc()`.  This allows execution of R code to provide the necessary DBI driver to support the connection.

You can test this behavior and create connection objects manually with:
```{r eval=FALSE}
cfg <- config::get(file = "./path/to/conn.yml")
do.call(dbConnect, cfg$pg)
do.call(dbConnect, cfg$mssql)
do.call(dbConnect, cfg$oracle)
```

Or you can use the config file and `dbtest` to execute tests against all of these databae connections with:

```{r eval=FALSE}
dbtest::test_databases("./path/to/conn.yml")
```

### DBI Connection

The most straightforward way to interactively use `dbtest` is to provide a DBI connection object directly to `dbtest::test_single_database`.  

```{r eval=FALSE}
con <- DBI::dbConnect(odbc::odbc(), "My DSN")
dbtest::test_single_database(con)
```
