---
title: "dbtest"
output: github_document
---

[![Build Status](https://travis-ci.org/rstudio/dbtest.svg?branch=master)](https://travis-ci.org/rstudio/dbtest)

# Overview

The package uses **testthat** to automate dbplyr SQL translation by running the tests against live database connections.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Install

```{r, eval = FALSE}
devtools::install_github("rstudio/dbtest")
```

# Setup

The `config` package is used to secure the credentials and to allow to easily point the tests to different servers. The file needs to be named `config.yml`.

Here is a sample of the structure:

```{bash, eval = FALSE}
default:
  mssql:
    Driver: "MS SQL Driver"
    server: "localhost"
    Schema: "default"
    UID: "rstudio"
    PWD: "rstudio"
    port: 10000
```


# Usage

Use the `test_database` command to run the tests.  The arguments are as follows:
 
- **databases** - A vector of one or more databases that will be tested.  The value should match the name should match the section inside the `config.yml` file.  In the exampel above, we would use `mssql`

- **configuration** - Defaults to `default`.  It refers to the main configuration selected that is in the `config.yml` file.  

Use a variable to retain the results.  The `test_database` function returns a `data.frame` with the results.

```{r, include = FALSE}
library(dbtest)
```


```{r}
library(dbtest)
results <- test_database("sqlite")
```

The resulting `data.frame` has the following variables:

- **file** - Name of the R script    
- **context**  
- **test** - It has the test's name.  At this time, the names are kept to the single function name that was the focus of the test     
- **nb**       
- **failed**   
- **skipped** 
- **error**    
- **warning**  
- **user**    
- **system**   
- **real** - Time elapsed for the test     
- **database** 
- **result** - Contains the actual error returned
- **call** - Contains the code used in the test     
- **res** - If the test either fails or has errors then the value will be **Failed**, otherwise it will be **Passed**

# Analysis

The package has a quick-n-easy report that takes the output from the `test_databases` output and creates an html based report.

```{r, eval = FALSE}
html_report(results)
```

Another way I found was to just run a quick visualization

```{r, fig.height = 14, fig.width = 3}
library(ggplot2) 

ggplot(data = results) +
  geom_raster(aes(x = database, y = test, fill = res))
```



